Xolotl: An Intuitive and Approachable Neuron \& Network Simulator in MATLAB

### 2018-08-13 paper/paper.tex
cleanup, trying to fix the NEURON mess

	Conductance-based models of neurons have been used extensively in computational neuroscience. However, working with these models continues to be a challenge due to their high dimensionality and large number of parameters, that hinders intuition of their dynamical features. Here, we present a neuron and network simulator built on a novel automatic type system that binds class templates written in "C++" to object-oriented code in "MATLAB". Our approach builds on the tradition of uniting the speed and power of languages like "C++" with the ease-of-use and feature-set of scientific programming languages like "MATLAB". Our software allows for the creation and manipulation of hierarchical models with components that are named and searchable, permitting intuitive high-level programmatic control over all parameters. Our simulator's architecture allows for the live manipulation of any parameter in any model, and for visualizing the effects of changing that parameter immediately. Our simulator is fully featured with hundreds of ion channel models from the electrophysiological literature, and can be extended to include arbitrary conductances, synapses, and mechanisms. Finally, our simulator is written in a modular fashion and has been released under a permissive free software license, enabling seamless integration into the workflow of many researchers.

@@

Nervous systems process and transmit information using electrically excitable membranes. Conductance-based models are a powerful biophysical simplification of an electrically excitable compartment in a neuron (\cite{hodgkinQuantitativeDescriptionMembrane1952}). Studies based on the Hodgkin-Huxley formalism now contribute significantly to mainstream research in small circuits (\cite{marderTheoryMotion1995, prinzComputationalApproachesNeuronal2010, prinzInsightsModelsRhythmic2006}). These models provide an approachable framework for understanding many salient principles of neuroscience, since they explicitly model cell membranes and ion channels as electrical components in a circuit. However, several challenges remain in working with biophysically-detailed conductance-based neuron models. First, these models can be high-dimensional with many nonlinear differential equations, each with several parameters. Second, many or all equations in these models can be strongly coupled through dynamical variables like the membrane potential.  In multi-compartment models of spatially extended neurons, membrane potentials in every compartment can be different, and are coupled to each other. Finally, the choice of programming language used to implement the model imposes tradeoffs in designing and using neuron and network simulators: simulators written in languages like "C++" or "FORTRAN" can integrate equations quickly, but often lack the ease-of-use and interoperability of those written in scientific programming languages like "Python", "Julia", or "MATLAB".

@@

To overcome these design limitations, we have developed a novel automatic type system, "cpplab", which binds "MATLAB" code to "C++" header files. This architecture automatically creates objects in "MATLAB" that represent the underlying object-oriented "C++" code, allowing the symbolic manipulation of "C++" objects in the "MATLAB" interface. In this paper, we introduce "xolotl", an implementation of the "cpplab" system specialized in integrating conductance-based neuron and network models. Models can be easily constructed from network building blocks in a few lines of "MATLAB" code using a hierarchical and intuitive syntax. Since models in the "MATLAB" workspace are automatically linked to models in the "C++" implementation, configuring these objects in "MATLAB" transparently configures the underlying "C++" objects. "xolotl" comes packaged with hundreds of components that can be used to assemble cells and networks;  has built-in visualization functions to inspect voltage time traces and activation functions; and a GUI for real-time manipulation of model parameters. The ease-of-use of the tools built into our software lends them to pedagogical applications and the rapid prototyping of models. Our software aims to simplify the investigation of the dynamics of conductance-based network and neuron models, facilitate collaborative modeling, and is intended to complement other tools being developed in the computational neuroscience community.

@@

where \( C_{i} \) and  \( V_{i} \) are the capacitance and membrane potential of compartment  \( i \) and \( I_{j} \) is the current due to ion channel population \( j \) given by

@@

This software is designed to be used from within "MATLAB", a scientific programming language common amongst neuroscientists and engineers for pedagogy and research. We aimed to make "xolotl" completely usable entirely from within "MATLAB", and fully scriptable. Models created using this simulator appear in the "MATLAB" workspace as native objects, and are thus fully compatible with the large library of toolboxes that "MATLAB" provides, allowing the software to be used as a component of other packages and tools. All parameters of models and activation functions of any channel can be inspected before running the model. We designed several features of the simulator to be easily extensible: adding custom conductances or synapse types is possible by calling functions that generate new "C++" files on-the-fly. Finally, "xolotl" fully auditable by design, with several tools to verify model and parameter integrity and aid in reproducibility.

@@

\paragraph{Object-oriented.}
"xolotl" is designed to mirror the nested and hierarchical structure of networks and neurons. Biological neuronal networks are made up of neurons that are connected to each other with synapses. Each of these neurons contains within it a set of conductances and synaptic currents that contribute to its electrical behavior. Intracellular mechanisms can act within the cell to modify and regulate conductances, synapses, or the dynamic properties of the cell. Similarly, a "xolotl" model can contain a set of compartments that can represent individual neurons. Each compartment can contain an arbitrary set of conductances, and compartments, conductances and synapses can contain mechanisms that affect anything in the model. All objects in the "xolotl" model tree (compartments, conductances, etc.) are genuine "MATLAB" objects with their own type, properties, and associated methods. This object-oriented programming paradigm naturally represents the hierarchical structure of biological networks that contain neurons that contain populations of ion channels, and makes constructing, integrating, and thinking about these models easier.

\paragraph{Automatic type system.}
To circumvent the tradeoff between high-performance but hard to use languages like "C++" and richly-featured but potentially slow languages like "MATLAB", we constructed an automatic type system that links object oriented code in "C++" to object oriented code in "MATLAB". This architecture lets us construct the core of the simulator in "C++", leveraging features of "C++" like pointers that are not readily available in "MATLAB". A rudimentary way to make this "C++" code useable in "MATLAB" would be to re-write that code in "MATLAB" so that "MATLAB" objects can be bound their "C++" implementations. However, this approach is cumbersome and inefficient since only features common between "MATLAB" and "C++" can be used. Instead, our automatic type system creates objects in "MATLAB" on-the-fly from "C++" class specifications, obviating the need to rewrite code in "MATLAB" while preserving a tight coupling between objects in the "MATLAB" workspace and their "C++" implementation. Crucially, this method makes developing new code much easier and simplifies the task of constructing complex frameworks that span these two languages.

\paragraph{Automatic hashing and compiling.}
Since every model requires a compiled binary to run, a potential stumbling block is the problem of unambiguously linking a model to a binary executable. "xolotl" solves this problem by hashing the "C++" header files of every component in the model recursively, allowing a model, no matter how complex, to be compactly represented by a short alphanumeric identifier (its "hash"). Compiled binaries are named using this hash, ensuring both that the correct binary is run to integrate the model, and that compilation occurs only as needed. This powerful feature enables the user experience to remain entirely within the "MATLAB" workspace, with compilation and selection of the correct binary occurring silently in the background.

@@

\paragraph{Inefficient tools for handling large networks.} "xolotl" was designed to work with small but complex networks and models, where every compartment and component is named, rather than numbered. It is more therefore suited towards simulating small and heterogeneous networks rather than large but homogenous networks. While the software can integrate large networks ($>1000$ compartments), other tools are presumably more suited to this task, offering more natural frameworks for dealing with a large number of identical units.

@@

In this section, we illustrate how "xolotl" can be used to generate a variety of models and simulate them. These examples have been chosen to illustrate various features of "xolotl", and are intended to serve as templates upon which researchers and educators can build.

@@

The axon of the giant squid contains contains a fast inactivating sodium conductance ("NaV"), a slower non-inactivating potassium conductance ("Kd"), and a passive leak current ("Leak"). Seminal work by Hodgkin and Huxley showed that depolarizations of the membrane, for example, by injecting current, could lead to an activation of the voltage-sensitive "NaV" channels, which led to a run away depolarization that was terminated by the inactivation of "NaV" channels and the activation of "Kd" channels that repolarized the membrane (\cite{hodgkinComponentsMembraneConductance1952, hodgkinMeasurementCurrentvoltageRelations1952}). As one of the simplest models of excitable neural membranes, the Hodgkin-Huxley model often serves as a the first model introduced in pedagogical literature (\cite{dayanTheoreticalNeuroscience2001, sterrattPrinciplesComputationalModelling2011, trappenbergFundamentalsComputationalNeuroscience2010}).

In this example, we demonstrate how to simulate the spiking activity of a Hodgkin-Huxley-like model, and how the tools built into our simulator make it easy to set up and integrate the model and gain insight into the underlying biophysical mechanisms. This simple model consists of a single-compartment neuron with three types of conductances (Figure \ref{fig:figurehh}A). This hierarchical organization of the neuron is mirrored in the structure of the model: an object of type "compartment" is used to represent the cell body, and three objects of type "conductance" are used to model the three populations of ion channels (Figure \ref{fig:figurehh}B). These models of ion channels were obtained from \cite{liuModelNeuronActivityDependent1998} based on electrophysiological recordings from the lobster stomatogastric ganglion (\cite{turrigianoSelectiveRegulationCurrent1995}), and are part of the simulator. The syntax of constructing the neuron model (using "add" statements) preserves this hierarchical organization, and the code snippet in (Figure \ref{fig:figurehh}C) shows how this model can be created using just a few lines of code.

Adding an injected current and calling the built-in "plot" function plots the time series of membrane voltage. In the absence of injected current, the model is quiescent. When 0.2 nA is injected, the model tonically spikes (Figure \ref{fig:figurehh}A-B).  The \texttt{plot} function displays a voltage trace colored by the dominant current (Figure \ref{fig:figurehh}A). Colors in the voltage trace indicate the strongest instantaneous inward current when the voltage is increasing, and strongest instantaneous outward current when the voltage is decreasing. These built-in features reveal that the dominant current during the upswing of every action potential is the sodium current, and the dominant current immediately after the peak of the action potential is the potassium current, but that the leak current contributes to depolarization following an action potential. This feature could be useful in quickly understanding the contributions of a number of ion channel types in a complex voltage trace from a more complicated neuron model.

@@

"xolotl" can reproduce such a voltage clamp experiment \textit{in-silico}. Figure \ref{fig:figureclamp}B illustrates how a simple model with a single compartment and a single ion channel type can be set up and clamped to a desired voltage.  Integrating the model yields the current required to clamp the cell at that voltage. Here, we use a delayed-rectifier potassium conductance (\cite{liuModelNeuronActivityDependent1998}) and simulate a voltage-clamp experiment whose goal is to infer the activation function of this channel. First, the cell is clamped to a number of different voltages (Fig. \ref{fig:figureclamp}C) and the resultant clamping currents are measured by integrating the model (Fig. \ref{fig:figureclamp}D). The asymptotic clamping currents depend on the clamping voltage in a nonlinear manner (Fig. \ref{fig:figureclamp}E), since the open probabilities of the channel are functions of the membrane voltage. Assuming the reversal potential is known, Eq. \eqref{eq:2} can be used to solve for the total conductance of the channel as a function of the clamping voltage (Fig. \ref{fig:figureclamp}F). Finally, a sigmoid can be fit to the normalized conductance-voltage curves to obtain the activation function of the ion channel population (Fig. \ref{fig:figureclamp}G-H). "xolotl" can therefore be used to describe graphically the theoretical underpinnings of ion channel characterization through voltage clamp and can serve as an effective pedagogical tool in computational neuroscience.

@@

So far, the models we described only considered the voltage dynamics of a cell (the solution to Eq. \ref{eq:1}). However, real neurons possess a several dynamical features, stemming from a number of intracellular mechanisms. "xolotl" makes it possible to model and include arbitrary intracellular mechanisms. In our simulator, these intracellular mechanisms are represented by the "mechanism" object, and can be bound to compartments, conductances, and other object types.

A key intracellular mechanisms is the cytosolic buffering of Calcium and its influx through voltage-gated Calcium channels. Figure \ref{fig:figuremechanism}A shows a model of a single-compartment model with 8 populations of ion channels (\cite{liuModelNeuronActivityDependent1998}). Without any mechanism for Calcium influx or buffering, the intracellular Calcium levels in this model do not change (Figure \ref{fig:figuremechanism}B) and the model tonically spikes (Figure \ref{fig:figuremechanism}C). Calcium buffering and influx can be modeled by a differential equation that increases intracellular Calcium with the current through Calcium channels and relaxes back to a baseline value (\cite{liuModelNeuronActivityDependent1998, prinzAlternativeHandtuningConductancebased2003, dayanTheoreticalNeuroscience2001}) (Figure \ref{fig:figuremechanism}D). This mechanism exists in the "xolotl" library as "CalciumMech1" and can be added to the model using a simple "add" statement (Figure \ref{fig:figuremechanism}E). The intracellular Calcium in the model now oscillates periodically (Figure \ref{fig:figuremechanism}F), synchronized to bursts in action potentials in this cell (Figure \ref{fig:figuremechanism}G).

Neurons can regulate their electrical activity by controlling the spectrum of ion channels they express (\cite{macleanActivityIndependentHomeostasisRhythmically2003, turrigianoSelectiveRegulationCurrent1995, schulzVariableChannelExpression2006}). Here, we will show how "xolotl" can be used to represent a recently proposed model of a homeostatic feedback system that controls the transcription rates of ion channels with the integral of an error signal derived from the intracellular Calcium concentration (\cite{olearyCorrelationsIonChannel2013, olearyCellTypesNetwork2014}) (Figure \ref{fig:figuremechanism}H). Since this mechanism affects each ion channel population individually, an object corresponding to this mechanism is added to each "conductance" object in the neuron (Figure \ref{fig:figuremechanism}I). Setting all maximal conductance densities to some low value and integrating the model shows that the intracellular Calcium levels rise over time and approach the target Calcium concentration (Figure \ref{fig:figuremechanism}J), while all conductance densities increase and then remain bounded (Figure \ref{fig:figuremechanism}K). Examining the voltage dynamics of the cell reveal that it transitions from quiescence to truncated bursts of action potentials to periodic bursting as this mechanism regulates the neuron's ion channel spectrum.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
%
%		SNAPSHOTS
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Switching back and forth points in parameter space and state space of a neuron model is a common occurrence in working with neuron models (\cite{deschutterConsumerGuideNeuronal1992}). "xolotl" makes it easy to bookmark configurations of a model and return to them at-will. Internally, "xolotl" does not distinguish between parameters and dynamical variables of a model, and uses the "serialize" method to gather all parameters and dynamic variables into a vector of values that is passed to the underlying "C++" implementation. A paired "deserialize" method is used to update all parameters and variables in the object tree from this vector. This architecture provides a natural framework for representing the state of any model, no matter how complex, using a vector of numbers. The "snapshot" method built into "xolotl" leverages this schema to save the entire state of the model in a named variable, that can be accessed using another built in method called "reset".

(Figure \ref{fig:figuresnapshot}) illustrates how these features can be used to explore the dynamics of the model presented in the previous section. First, the current state of the model is saved using the "snapshot" method into a namespace called ``initial''. In this state, the model exhibits periodic bursting activity due to a particular configuration of maximal conductance densities (Figure \ref{fig:figuremechanism}, orange).  On setting the maximal conductances of the Calcium-permissive channels to 0, the model switches to a tonic spiking activity (Figure \ref{fig:figuremechanism}, purple). Integrating the model for a longer duration allows the homeostatic control mechanism in the cell to restore the conductance profile and bursting activity to a state close to the original state (Figure \ref{fig:figuremechanism}, green). This state is saved using the name ``final''.

The initial state can be returned to using the "reset" method, and a new manipulation to the model can be explored. Here, the intracellular Calcium target is modified, and the model is re-integrated, to yield a different voltage activity (Figure \ref{fig:figuremechanism}, blue). At the end of this numerical exploration, any of the saved states can be quickly returned to using the "reset" method, making the process of re-initializing models to desired states and parameters both error-free and painless.

@@

Neurons communicate via synapses, electrochemical junctions between cells (\cite{gjorgjievaComputationalImplicationsBiophysical2016, huaNeuralActivityDynamics2004}). In chemical synapses, the presynaptic neuron releases neurotransmitter across the synaptic cleft, which activate receptors on the postsynaptic neuron. In electrical synapses, no chemical intermediary is involved. Ions flux directly from the presynaptic neuron to the postsynaptic one. New patterns of activity emerge from the characteristics of the connecting synapses (\cite{liComodulationSynapsesVoltagegated2018, nadimSynapticDepressionCreates1999, gutierrezRectifyingElectricalSynapses2013, gutierrezMultipleMechanismsSwitch2013}).

Network models in "xolotl" consist of "compartment" objects connected by "synapse" objects. Synapses are stored in a vector array as a field of the "xolotl" object in "MATLAB". Presynaptic and postsynaptic labels indicate the connectivity of the synapse. Like conductances, the dynamics of the synapse is specified by the type of "cpplab" object added to the network structure. For example, the "Electrical" synapse type describes equipotent and rectifying electrical synapses.

Postsynaptic currents are a function of the membrane potential of both the presynaptic and postsynaptic neurons. "xolotl" handles synaptic current equations in the form

\begin{equation}
	I_{syn} = \bar{g}_{syn} s(V_{pre}) (V_{post} - E_{syn})
\end{equation}

Here $I_{syn}$ is the postsynaptic current, $\bar{g}_{syn}$ is the maximal conductance of the synaptic connection, $s$ is a dynamical variable dependent on the presynaptic membrane potential. The electrochemical driving force is comprised of the postsynaptic membrane potential and the reversal potential of the synapse.

>>>>

	Conductance-based models of neurons have been used extensively in computational neuroscience. However, working with these models continues to be a challenge due to their high dimensionality and large number of parameters. Here, we present a neuron and network simulator built on a novel automatic type system that binds object-oriented code written in "C++" to objects in "MATLAB". Our approach builds on the tradition of uniting the speed and power of languages like "C++" with the ease-of-use and feature-set of scientific programming languages like "MATLAB". Our software allows for the creation and manipulation of hierarchical models with components that are named and searchable, permitting intuitive high-level programmatic control over all parameters. Our simulator's architecture allows for the live manipulation of any parameter in any model, and for visualizing the effects of changing that parameter immediately. Our simulator is fully featured with hundreds of ion channel models from the electrophysiological literature, and can be extended to include arbitrary conductances, synapses, and mechanisms. Finally, our simulator is written in a modular fashion and has been released under a permissive free software license, enabling seamless integration into the workflow of other researchers.

@@

Nervous systems process and transmit information using electrically excitable membranes. Conductance-based models are a powerful biophysical simplification of an electrically excitable compartment in a neuron (\cite{hodgkinQuantitativeDescriptionMembrane1952}). Studies based on the Hodgkin-Huxley formalism now contribute significantly to mainstream research in small circuits (\cite{marderTheoryMotion1995, prinzComputationalApproachesNeuronal2010, prinzInsightsModelsRhythmic2006}). These models provide an approachable framework for understanding many salient principles of electrophysiology, since they explicitly model cell membranes and ion channels as electrical components in a circuit. However, several challenges remain in working with biophysically-detailed conductance-based neuron models. First, these models can be high-dimensional with many nonlinear differential equations, each with several parameters. Second, many or all equations in these models can be strongly coupled through dynamical variables like the membrane potential.  In multi-compartment models of spatially extended neurons, membrane potentials in every compartment can be different, and are coupled to each other. Finally, the choice of programming language used to implement the model imposes tradeoffs in designing and using neuron and network simulators: simulators written in languages like "C++" or "FORTRAN" can integrate equations quickly, but often lack the ease-of-use and interoperability of those written in scientific programming languages like "Python", "Julia", or "MATLAB".

@@

To overcome these design limitations, we have developed a novel automatic type system, that we call "cpplab", which binds "MATLAB" code to classes specified in "C++" header files. This architecture automatically creates objects in "MATLAB" that represent the underlying object-oriented "C++" code, allowing the symbolic manipulation of "C++" objects in the "MATLAB" interface. In this paper, we introduce "xolotl", an implementation of the "cpplab" system specialized in integrating conductance-based neuron and network models. Models can be easily constructed from network building blocks in a few lines of "MATLAB" code using a hierarchical and intuitive syntax. Since models in the "MATLAB" workspace are automatically linked to models in the "C++" implementation, configuring these objects in "MATLAB" transparently configures the underlying "C++" objects. "xolotl" comes packaged with hundreds of components that can be used to assemble cells and networks;  has built-in visualization functions to inspect voltage time traces and activation functions; and a GUI for real-time manipulation of model parameters. The ease-of-use of the tools built into our software lends them to pedagogical applications and the rapid prototyping of models. Our software aims to simplify the investigation of the dynamics of conductance-based network and neuron models, facilitate collaborative modeling, and is intended to complement other tools being developed in the computational neuroscience community.

@@

where \( C_{i} \) and  \( V_{i} \) are the capacitance and membrane potential of compartment  \( i \). Compartments can represent whole neurons or parts of neurons.  \( I_{j} \) is the current due to ion channel population \( j \) and is given by

@@

This software is designed to be used from within "MATLAB", a scientific programming language common amongst neuroscientists and engineers for pedagogy and research. Our goal was to make "xolotl" completely usable entirely from within "MATLAB". Models created using this simulator appear in the "MATLAB" workspace as native objects, are are thus fully scriptable, and are fully compatible with the large library of toolboxes that "MATLAB" provides, allowing the software to be used as a component of other packages and tools. All parameters of models and activation functions of any channel can be inspected at any point. We designed several features of the simulator to be easily extensible: adding custom conductances or synapse types is possible by calling functions that generate new "C++" files on-the-fly. Finally, "xolotl" fully auditable by design, with several tools to verify model and parameter integrity and aid in reproducibility.

@@

\paragraph{Object-oriented.}
"xolotl" is designed to mirror the nested and hierarchical structure of networks and neurons. Biological neuronal networks are made up of neurons that are connected to each other with synapses. Each of these neurons contains within it a set of conductances and synaptic currents that contribute to its electrical behavior. Intracellular mechanisms can act within the cell, or parts of the cell, to modify and regulate conductances, synapses, or other dynamic properties of the cell. Similarly, a "xolotl" model can contain a set of compartments that can represent individual neurons. Each compartment can contain an arbitrary set of conductances. Compartments, conductances and synapses can contain mechanisms that can affect anything in the model. All objects in the "xolotl" model tree (compartments, conductances, etc.) are bonafide "MATLAB" objects with their own type and properties. This object-oriented programming paradigm naturally represents the hierarchical structure of biological networks that contain neurons that contain populations of ion channels, and makes constructing, integrating, and thinking about these models easier.

\paragraph{Automatic type system.}
To circumvent the tradeoff between high-performance but hard to use languages like "C++" and richly-featured but potentially slow languages like "MATLAB", we constructed an automatic type system that links object oriented code in "C++" to object oriented code in "MATLAB". This architecture lets us construct the core of the simulator in "C++", leveraging features of "C++" like pointers that are not readily available in "MATLAB". A rudimentary way to make this "C++" code useable in "MATLAB" would be to re-write that code in "MATLAB" so that "MATLAB" objects can be bound to their "C++" implementations. However, this approach is cumbersome and inefficient, and can introduce errors. Instead, our automatic type system creates objects in "MATLAB" on-the-fly from "C++" class specifications, obviating the need to rewrite code in "MATLAB" while preserving a tight coupling between objects in the "MATLAB" workspace and their "C++" implementation. Crucially, this method makes developing new code much easier and simplifies the task of constructing complex frameworks that span these two languages.

\paragraph{Automatic hashing and compiling.}
Since every model requires a compiled binary to run, a potential stumbling block is the problem of unambiguously linking a model to a binary executable. "xolotl" solves this problem by hashing (\cite{rivestMD5MessageDigestAlgorithm1992}) the "C++" header files of every component in the model recursively, allowing a model, no matter how complex, to be compactly represented by a short alphanumeric identifier (its ``hash''). Compiled binaries are named using this hash, ensuring both that the correct binary is run to integrate the model, and that compilation occurs only as needed. This powerful feature enables the user experience to remain entirely within the "MATLAB" workspace, with compilation and selection of the correct binary occurring silently in the background.

@@

\paragraph{Inefficient tools for handling large networks.} "xolotl" was designed to work with small but complex networks and models, where every compartment and component is named, rather than numbered. It is more therefore suited towards simulating small and heterogeneous networks rather than large but homogenous networks. While the software can integrate large networks ($>1000$ compartments), other tools are presumably more suited to this task, offering more natural frameworks for dealing with a large number of identical units. Similarly, "xolotl" is not optimized to solve coupled ODEs on complex branched morphologies, and other simulators like (\cite{hinesNEURONSimulationEnvironment1997}) are more suited for this task.

@@

In this section, we illustrate how "xolotl" can be used to generate, inspect, and simulate a variety of models. These examples have been chosen to illustrate various features of "xolotl", and are intended to serve as templates upon which researchers and educators can build.

@@

The axon of the giant squid contains contains a fast inactivating sodium conductance ("NaV"), a slower non-inactivating potassium conductance ("Kd"), and a passive leak current ("Leak"). Seminal work by Hodgkin and Huxley showed that depolarizations of the membrane could lead to an activation of the voltage-sensitive "NaV" channels, which led to a run away depolarization that was terminated by the inactivation of "NaV" channels and the activation of "Kd" channels that repolarized the membrane (\cite{hodgkinComponentsMembraneConductance1952, hodgkinMeasurementCurrentvoltageRelations1952}). As one of the simplest models of excitable neural membranes, the Hodgkin-Huxley model often serves as a the first model introduced in pedagogical literature (\cite{dayanTheoreticalNeuroscience2001, sterrattPrinciplesComputationalModelling2011, trappenbergFundamentalsComputationalNeuroscience2010}).

In this example, we demonstrate how to simulate the spiking activity of a Hodgkin-Huxley-like model, and how the tools built into our simulator make it easy to set up and integrate the model and gain insight into the underlying biophysical mechanisms. This simple model consists of a single electrical compartment with three types of conductances (Figure \ref{fig:figurehh}A). This hierarchical organization of the neuron is mirrored in the structure of the model in our simulator: an object of type "compartment" is used to represent the cell body, and three objects of type "conductance" are used to model the three populations of ion channels (Figure \ref{fig:figurehh}B). These models of ion channels were obtained from \cite{liuModelNeuronActivityDependent1998} based on electrophysiological recordings from the lobster stomatogastric ganglion (\cite{turrigianoSelectiveRegulationCurrent1995}), and are part of the simulator. The code to set up this model is terse, idiomatic, relies on no special markup, and preserves the heirarchical nature of the model (Figure \ref{fig:figurehh}C).

Adding an injected current and calling the built-in "plot" function plots the time series of membrane voltage. In the absence of injected current, the model is quiescent. When 0.2 nA is injected, the model tonically spikes (Figure \ref{fig:figurehh}A-B).  The \texttt{plot} function displays a voltage trace colored by the dominant current (Figure \ref{fig:figurehh}A). Colors in the voltage trace indicate the strongest instantaneous inward current when the voltage is increasing, and strongest instantaneous outward current when the voltage is decreasing. This built-in feature reveals that the dominant current during the upswing of every action potential is the sodium current, and the dominant current immediately after the peak of the action potential is the potassium current, but that the leak current contributes to depolarization following an action potential. This feature could be useful in quickly understanding the contributions of a number of ion channel types in a complex voltage trace from a more complicated neuron model.

@@

"xolotl" can reproduce such a voltage clamp experiment \textit{in-silico}. Figure \ref{fig:figureclamp}B illustrates how a simple model with a single compartment and a single ion channel type can be set up and clamped to a desired voltage.  Integrating the model yields the current required to clamp the cell at that voltage. Here, we use a delayed-rectifier potassium conductance (\cite{liuModelNeuronActivityDependent1998}) and simulate a voltage-clamp experiment whose goal is to infer the activation function of this channel. First, the cell is clamped to a number of different voltages (Fig. \ref{fig:figureclamp}C) and the resultant clamping currents are measured by integrating the model (Fig. \ref{fig:figureclamp}D). Since the compartment is being voltage clamped, integrating the model returns the clamping current:

\begin{lstlisting}[style=Matlab-editor]
I_clamp = x.integrate;
\end{lstlisting}

By repeating the integration at a number of clamping voltages, we observe that the asymptotic clamping currents depend on the clamping voltage in a nonlinear manner (Fig. \ref{fig:figureclamp}E), since the open probabilities of the channel are functions of the membrane voltage. Assuming the reversal potential is known, Eq. \eqref{eq:2} can be used to solve for the total conductance of the channel as a function of the clamping voltage (Fig. \ref{fig:figureclamp}F). Finally, a sigmoid can be fit to the normalized conductance-voltage curves to obtain the activation function of the ion channel population (Fig. \ref{fig:figureclamp}G-H). "xolotl" can therefore be used to describe graphically the theoretical underpinnings of ion channel characterization through voltage clamp and can serve as an effective pedagogical tool in computational neuroscience.

@@

So far, the models we described only considered the voltage dynamics of a cell (the solution to Eq. \ref{eq:1}). However, real neurons possess several dynamical features, arising from a variety of intracellular mechanisms. "xolotl" makes it possible to model and include arbitrary intracellular mechanisms. In our simulator, these intracellular mechanisms are represented by the "mechanism" object, and can be bound to compartments, conductances, and other object types.

A key intracellular mechanisms is the cytosolic buffering of Calcium and its influx through voltage-gated Calcium channels. Figure \ref{fig:figuremechanism}A shows a model of a single-compartment model with 8 populations of ion channels (\cite{liuModelNeuronActivityDependent1998}). Without any explicit mechanism for Calcium influx or buffering, the intracellular Calcium levels in this model do not change (Figure \ref{fig:figuremechanism}B) and the model tonically spikes (Figure \ref{fig:figuremechanism}C). Calcium buffering and influx can be modelled by a differential equation that increases intracellular Calcium with the current through Calcium channels and relaxes back to a baseline value (\cite{liuModelNeuronActivityDependent1998, prinzAlternativeHandtuningConductancebased2003, dayanTheoreticalNeuroscience2001}) (Figure \ref{fig:figuremechanism}D). This mechanism exists in the "xolotl" library as "CalciumMech1" and can be added to the model using a simple "add" statement (Figure \ref{fig:figuremechanism}E). The intracellular Calcium in the model now oscillates periodically (Figure \ref{fig:figuremechanism}F), synchronized to bursts in action potentials in this cell (Figure \ref{fig:figuremechanism}G).

Neurons can regulate their electrical activity by controlling the spectrum of ion channels they express (\cite{macleanActivityIndependentHomeostasisRhythmically2003, turrigianoSelectiveRegulationCurrent1995, schulzVariableChannelExpression2006}). Here, we will show how "xolotl" can be used to represent a recently proposed model of a homeostatic feedback system that controls the transcription rates of ion channels with the integral of an error signal derived from the intracellular Calcium concentration (\cite{olearyCorrelationsIonChannel2013, olearyCellTypesNetwork2014}) (Figure \ref{fig:figuremechanism}H). Since this mechanism affects each ion channel population individually, an object corresponding to this mechanism is added to each "conductance" object in the neuron (Figure \ref{fig:figuremechanism}I). Setting all maximal conductance densities to some low value and integrating the model shows that the intracellular Calcium levels rise over time and approach the target Calcium concentration (Figure \ref{fig:figuremechanism}J), while all conductance densities increase and then remain bounded (Figure \ref{fig:figuremechanism}K). Examining the voltage dynamics of the cell reveal that it transitions from quiescence to truncated bursts of action potentials to periodic bursting as this mechanism regulates the neuron's ion channel spectrum. In summary, our simulator can be used to construct neuron models with intracellular mechanisms such as Calcium influx and buffering, and homeostatic regulation.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
%
%		SNAPSHOTS
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Switching back and forth points in parameter space and state space of a neuron model is a common occurrence in working with neuron models, and a significant fraction of a modeller's time is spent in a feedback loop of running simulations, viewing the output, changing parameters, and re-running simulations. (\cite{deschutterConsumerGuideNeuronal1992}). "xolotl" makes it easy to bookmark configurations of a model and return to them at will. Internally, "xolotl" uses the "serialize" method to gather all parameters and dynamic variables into a vector of values that is passed to the underlying "C++" implementation. A paired "deserialize" method is used to update all parameters and variables in the object tree from this vector. This architecture provides a natural framework for representing the state of any model, no matter how complex, using a vector of numbers. The "snapshot" method built into "xolotl" leverages this schema to save the entire state of the model in a named variable, that can be accessed using another built-in method called "reset".

(Figure \ref{fig:figuresnapshot}) illustrates how these features can be used to explore the dynamics of the model presented in the previous section. First, the current state of the model is saved using the "snapshot" method into a state called ``initial''. In this state, the model exhibits periodic bursting activity due to a particular configuration of maximal conductance densities (Figure \ref{fig:figuremechanism}, orange).  On setting the maximal conductances of the Calcium-permissive channels to 0, the model switches to a tonic spiking activity (Figure \ref{fig:figuremechanism}, purple). Integrating the model for a longer duration allows the homeostatic control mechanism in the cell to restore the conductance profile and bursting activity to a state close to the original state (Figure \ref{fig:figuremechanism}, green). This state is saved using the name ``final''.

The initial state can be returned to using the "reset" method, and a new manipulation to the model can be explored. Here, the intracellular Calcium target is modified, and the model is re-integrated, to yield a different voltage activity (Figure \ref{fig:figuremechanism}, blue). At the end of this numerical exploration, any of the saved states can be quickly returned to using the "reset" method, making the process of re-initializing models to desired states and parameters both error-free and efficient.

@@

Neurons communicate and interact using synapses, electrochemical junctions between cells (\cite{gjorgjievaComputationalImplicationsBiophysical2016, huaNeuralActivityDynamics2004}). In chemical synapses, the presynaptic neuron releases packets of neurotransmitter across the synaptic cleft, which activate receptors on the postsynaptic neuron. In electrical synapses, no chemical intermediary is involved. Ions flux directly from the presynaptic neuron to the postsynaptic one, setting up a current in the post-synaptic cell. New patterns of activity can emerge from the characteristics of the connecting synapses (\cite{liComodulationSynapsesVoltagegated2018, nadimSynapticDepressionCreates1999, gutierrezRectifyingElectricalSynapses2013, gutierrezMultipleMechanismsSwitch2013}).

Network models in "xolotl" consist of "compartment" objects that can be connected by "synapse" objects. Two compartments representing different neurons can be connected using synapses using the built-in "connect" method. For example, to connect two single-compartment neurons called "LP" and "PY" using a chemical synapse of type "Cholinergic" with strength 30 nS, we can use

\begin{lstlisting}[style=Matlab-editor]
x.connect('LP','PY','Cholinergic','gbar',30); 
\end{lstlisting}

Our simulator has several types of synapses like Cholinergic, Glutamatergic, AMPA and NMDA built-in, and other synapse classes can be easily added using templates. Figure \ref{fig:figurenetwork} demonstrates the implementation of a model of the triphasic pyloric rhythm in the stomatogastric ganglion of crustaceans (\cite{prinzSimilarNetworkActivity2004}). The pyloric model contains three compartments ("AB", "LP", and "PY") and seven synapses (Figure \ref{fig:figurenetwork}A). This structure is recapitulated in the hierarchy of the "xolotl" object, where conductances are contained within compartments (Figure \ref{fig:figurenetwork}B). The membrane potentials show triphasic rhythmicity in the three compartments (Figure \ref{fig:figurenetwork}C-E). When the "PY" is depolarized, the dynamical variable mediating the  glutamatergic ("Glut") synapse model between "PY" and "LP" is close to 1 and "LP" is inhibited (Figure \ref{fig:figurenetwork}F). Conversely, when "PY" is hyperpolarized, the dynamical variable is close to 0. When "PY" spikes, IPSPs can be seen in the "LP" voltage trace (Figure \ref{fig:figurenetwork}D-E).
